Racial and Ethnic Differences in ETI: Outcomes

### 2024-05-17 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti_outcomes.qmd
separate scripts for mcp and segmented

```{r data setup}
#| include: false
# Import
encounter <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_encountersMerged_Del1.csv", na.strings = "")
annual <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_Annualized_Del1.csv", na.strings = "")
demo <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_DemogCFDiag_Del1.csv", na.strings = "")
# Race variable
races <- list(
  "Race1" = "White",
  "Race2" = "Black or African American",
  "Race3" = "American Indian or Alaska Native",
  "Race4" = "Asian",
  "Race5" = "Native Hawaiian or Other Pacific Islander",
  "Race6" = "Some other race"
)
demo$Race <- apply(demo[, paste0("Race", 1:6)], 1, function(r) {
  w <- which(r == 1)
  if (length(w) > 1 | 6 %in% w) {
    return("Mixed/Other/Unknown Race")
  } else {
    return(as.character(races[w]))
  }
})
demo$Race[demo$Hispanicrace == 1] <- "Hispanic or Latino"
demo$Race[is.na(demo$Hispanicrace)] <- "Mixed/Other/Unknown Race"
demo$Race <- factor(demo$Race,
  levels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "American Indian or Alaska Native",
    "Native Hawaiian or Other Pacific Islander", "Mixed/Other/Unknown Race"
  ),
  labels = c(
    "Non-Hispanic White", "Black or African American", "Hispanic or Latino", "Asian",
    "Other", "Other", "Other"
  )
)
# Insurance
ins_vars <- c(paste0("insurance", 1:7), "noinsurance")
ins <- c(
  "Health Insurance Policy (e.g. Private Insurance)", "Medicare", "Medicaid",
  "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
  "TriCare or other military health plan",
  "Indian Health Service", "Other", "No Insurance"
)
annual$Insurance <- apply(annual[, ins_vars], 1, function(r) {
  w <- which(r == 1)
  if (length(w) == 0) {
    return(NA)
  } else if (length(w) > 1) {
    return("More than one")
  } else {
    return(ins[w])
  }
})
annual$Insurance <- factor(annual$Insurance,
  levels = c(
    "Health Insurance Policy (e.g. Private Insurance)",
    "Indian Health Service",
    "Medicaid",
    "Medicare",
    "More than one",
    "No Insurance",
    "Other",
    "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
    "TriCare or other military health plan"
  ),
  labels = c(
    "Private", "Public", "Public", "Public", "More Than One", "No Insurance",
    "No Insurance", "Public", "Tricare"
  )
)
# Merge
encounter <- left_join(
  encounter,
  demo %>% select(eDWID, Race, Modulator_trikafta_first_date)
)
encounter <- left_join(encounter,
  annual %>% select(eDWID, ReviewYear, Insurance),
  by = join_by(eDWID, reviewyear == ReviewYear)
)
# Age group
encounter$age_group <- cut(encounter$encounterage, breaks = c(6, 12, 18, 30, Inf), right = F)
# Days between encounter and first ETI
encounter$Modulator_trikafta_first_date <-
  mdy(encounter$Modulator_trikafta_first_date)
encounter$encounterdate <- mdy(encounter$encounterdate)
encounter$Days <- as.numeric(encounter$encounterdate - encounter$Modulator_trikafta_first_date)
# Separate cleaned set
cleaned <- encounter %>% filter(gli_fev1_ppred_rn <= 200, !is.na(Days))
# List of outcomes
continuous_outcomes <- c(
  "gli_fev1_ppred_rn", "gli_fvc_ppred_rn", "gli_fev1fvc_ppred_rn",
  "bmipercentile", "heightpercentile", "weightpercentile"
)
binary_outcomes <- c(
  "staphylococcus_aureus", "haemophilus_influenzae", "pseudomonasaeruginosa",
  "burkho_complex"
)
# Save dataset for mcp model script
save(cleaned, file = "./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/mcp_dataset.RData")
```

```{r functions}
# Plots
loess_plot <- function(df = cleaned, outcome, facet = F) {
  p <- ggplot(df, aes(x = Days, y = !!sym(outcome), color = Race)) +
    geom_point(
      shape = ".", aes(x = Days, y = !!sym(outcome)), inherit.aes = F,
      alpha = 0.1
    ) +
    geom_smooth() +
    theme_classic()
  if (facet) {
    p <- p + facet_wrap(~Race)
  }
  return(p)
}
# Basic segmented model
basic_segmented <- function(df = cleaned, outcome, n_change_points = NULL,
                            change_points = NULL) {
  # Formula
  f <- as.formula(paste0(outcome, "~ Days+Race + age_group + Insurance"))
  # Basic linear model
  lmod <- lm(f, data = df)
  # Segmented model
  lmod_seg <- segmented(lmod, seg.Z = ~Days, psi = 0, control = seg.control(it.max = 0))
  # Plot
  d <- lmod$model
  d$pred <- fitted(lmod_seg)
}
```

@@

1. For now, I have excluded any rows where the FEV1 predicted values are over 200.

@@

The following models are adjusted for age group and insurance status.

@@

## Frequentist models

First, a Gaussian mixed model is fit with random intercept for patient ID and random slope for each patient. Then, the `segmented.lme` function is applied to this mixed model, with 0 as the starting value for finding the optimal breakpoint. The changepoint is also treated as a random effect and allowed to vary within a person.

```{r}
#| cache: true
#| eval: false
# Fit initial model
ri_rs_mod_fev1 <- lme(gli_fev1_ppred_rn ~ Race * Days + age_group + Insurance,
  random = ~ 1 + Days | eDWID, data = cleaned, na.action = na.omit
)
# Find change point
ri_rs_mod_seg_fev <- segmented.lme(ri_rs_mod_fev1, ~Days, psi = 0,
  random = list(eDWID = pdDiag(~ 1 + Days + U + G0)), # ind. random eff. (in the changepoint too)
  control = seg.control(n.boot = 0, display = F)
)
```

```{r}

# Questions for Scott and Kam

1. Some of the predicted lung function measures are insanely high. How should we deal with outliers? Is there a limit that you would consider biologically impossible? Over 200%? What about FVC and FEV1/FVC?

2. Similarly, what would be a good cutoff for the lower end? I assume that a value of 0 is not possible.

3. I know we've discussed this before, but looking at the LOESS plots, do we want to cut off the pre-ETI at about 2000 days?

4. Are we interested in statistical inference on age group? Or are we simply adjusting for it? If we're not clinically interested in the age groups is would be worth treating it as a continuous variable.

5. Do the linearity assumptions seem accurate to you?

6. Why do you think the BMI, height, etc. percentiles look so strange? Do they normally look like this?

```{r}
#| warning: false
loess_plot(df = encounter, outcome = "gli_fev1_ppred_rn")
```

- Is the second change point due to who is being collected after one year out?
  - 
- Kam will

< 20% predicted and >=150 (Kam)

>>>>

```{r data setup}
#| include: false
# Import
encounter <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_encountersMerged_Del1.csv", na.strings = "")
annual <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_Annualized_Del1.csv", na.strings = "")
demo <- read.csv("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Raw/DataDelivery_20240125/CFF22_DemogCFDiag_Del1.csv", na.strings = "")
# Race variable
races <- list(
  "Race1" = "White",
  "Race2" = "Black or African American",
  "Race3" = "American Indian or Alaska Native",
  "Race4" = "Asian",
  "Race5" = "Native Hawaiian or Other Pacific Islander",
  "Race6" = "Some other race"
)
demo$Race <- apply(demo[, paste0("Race", 1:6)], 1, function(r) {
  w <- which(r == 1)
  if (length(w) > 1 | 6 %in% w) {
    return("Mixed/Other/Unknown Race")
  } else {
    return(as.character(races[w]))
  }
})
demo$Race[demo$Hispanicrace == 1] <- "Hispanic or Latino"
demo$Race[is.na(demo$Hispanicrace)] <- "Mixed/Other/Unknown Race"
demo$Race <- factor(demo$Race,
  levels = c(
    "White", "Black or African American", "Hispanic or Latino", "Asian",
    "American Indian or Alaska Native",
    "Native Hawaiian or Other Pacific Islander", "Mixed/Other/Unknown Race"
  ),
  labels = c(
    "Non-Hispanic White", "Black or African American", "Hispanic or Latino", "Asian",
    "Other", "Other", "Other"
  )
)
# Insurance
ins_vars <- c(paste0("insurance", 1:7), "noinsurance")
ins <- c(
  "Health Insurance Policy (e.g. Private Insurance)", "Medicare", "Medicaid",
  "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
  "TriCare or other military health plan",
  "Indian Health Service", "Other", "No Insurance"
)
annual$Insurance <- apply(annual[, ins_vars], 1, function(r) {
  w <- which(r == 1)
  if (length(w) == 0) {
    return(NA)
  } else if (length(w) > 1) {
    return("More than one")
  } else {
    return(ins[w])
  }
})
annual$Insurance <- factor(annual$Insurance,
  levels = c(
    "Health Insurance Policy (e.g. Private Insurance)",
    "Indian Health Service",
    "Medicaid",
    "Medicare",
    "More than one",
    "No Insurance",
    "Other",
    "State special needs program, e.g., BCMH, CCS, CRS, GHPP, etc.",
    "TriCare or other military health plan"
  ),
  labels = c(
    "Private", "Public", "Public", "Public", "More Than One", "No Insurance",
    "No Insurance", "Public", "Tricare"
  )
)
# Merge
encounter <- left_join(
  encounter,
  demo %>% select(eDWID, Race, Modulator_trikafta_first_date)
)
encounter <- left_join(encounter,
  annual %>% select(eDWID, ReviewYear, Insurance),
  by = join_by(eDWID, reviewyear == ReviewYear)
)
# Age group
encounter$age_group <- cut(encounter$encounterage, breaks = c(6, 12, 18, 30, Inf), right = F)
# Days between encounter and first ETI
encounter$Modulator_trikafta_first_date <-
  mdy(encounter$Modulator_trikafta_first_date)
encounter$encounterdate <- mdy(encounter$encounterdate)
encounter$Days <- as.numeric(encounter$encounterdate - encounter$Modulator_trikafta_first_date)
# Remove impossible lung function values
encounter$gli_fev1_ppred_rn[encounter$gli_fev1_ppred_rn < 20 | encounter$gli_fev1_ppred_rn > 150] <- NA
encounter$gli_fvc_ppred_rn[encounter$gli_fvc_ppred_rn < 20 | encounter$gli_fvc_ppred_rn > 150] <- NA
encounter$gli_fev1fvc_ppred_rn[encounter$gli_fev1fvc_ppred_rn < 20 | encounter$gli_fev1fvc_ppred_rn > 150] <- NA
# Separate cleaned set
encounter <- encounter %>% filter(!is.na(Days))
# List of outcomes
continuous_outcomes <- c(
  "gli_fev1_ppred_rn", "gli_fvc_ppred_rn", "gli_fev1fvc_ppred_rn",
  "bmipercentile", "heightpercentile", "weightpercentile"
)
binary_outcomes <- c(
  "staphylococcus_aureus", "haemophilus_influenzae", "pseudomonasaeruginosa",
  "burkho_complex"
)
# Save dataset for mcp model script
save(encounter, file = "./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/cleaned_outcomes_dataset.RData")

```{r functions}
# Plots
loess_plot <- function(df = encounter, outcome, facet = F) {
  p <- ggplot(df, aes(x = Days, y = !!sym(outcome), color = Race)) +
    geom_point(
      shape = ".", aes(x = Days, y = !!sym(outcome)), inherit.aes = F,
      alpha = 0.1
    ) +
    geom_smooth() +
    theme_classic()
  if (facet) {
    p <- p + facet_wrap(~Race)
  }
  return(p)
}
```

@@

1. Lung function percent predicted values < 20% or > 150% were excluded.

@@

The following models are adjusted for age group and insurance status. First, a Gaussian mixed model is fit with random intercept for patient ID. Then, the `segmented.lme` function is applied to this mixed model, with no pre-specified starting values and `npsi = 2`. The change points are treated as random effects and allowed to vary within a person.

@@

```{r}
#| eval: false
load("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/fev1_single_cp_fit.RData")
load("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/fev1_double_cp_fit.RData")
```

- Is the second change point due to who is being collected after one year out?
